A memory frontier for complex synapses

### 2013-05-30 NIPS/SynapseNIPS.tex
nothing
almost final version

% -*- TeX -*- -*- US -*-
\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figs/"}}
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%
%additional symbols:
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A memory frontier for complex synapses}
%
% Author List:
%
\author{Subhaneil Lahiri and Surya Ganguli\\
Applied Physics Department, Stanford University, Stanford CA\\
\emaillink{sulahiri@stanford.edu}, \emaillink{sulahiri@stanford.edu}
%
}

@@

\begin{abstract}
  Blah blah blah.
\end{abstract}

@@

It is widely thought that our very ability to remember the past over long time scales depends crucially on our ability to modify synapses in our brain in an experience dependent manner.
Classical models of synaptic plasticity model synaptic efficacy as an analog scalar value, denoting the size of a post-synaptic potential injected into one neuron from another.
Theoretical work has shown that such models have a reasonable, extensive memory capacity, in which the number of long term associations that can be stored by a neuron is proportional its number of afferent synapses \cite{Hopfield1982model,Amit1985hopfield,Gardner1988perceptron}.
However, recent experimental work has shown that many synapses are more digital than analog; they cannot robustly assume an infinite continuum of analog values, but rather can only take on a finite number of distinguishable strengths, a number than can be as small as two \cite{Bliss1993LTP,Petersen1998allornone,O'Connor2005switch} (though see \cite{Enoki2009graded}).
This one simple modification leads to a catastrophe in memory capacity:  classical models with digital synapses, when operating in a palimpset mode in which the ongoing storage of new memories can overwrite previous memories, have a memory capacity proportional to the logarithm of the number of synapses \cite{amit1992constraints,amit1994learning}.
Intuitively, when synapses are digital, the storage of a new memory can flip a population of synaptic switches, thereby rapidly erasing previous memories. This result indicates that the dominant theoretical basis for the storage of long term memories in modifiable synaptic switches is flawed.

@@

Theoretical studies of complex, metaplastic synapses, have focused on analyzing the memory performance of a limited number of very specific, molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \autoref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely to be the case that there is a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

A simple mean memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ in those synapses that are potentiated and $\eq\M\dep$ in those synapses that are depressed.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{equation}\label{eq:evol}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{equation}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \e^{rt\frg} \w,
\end{equation}
%
where the weight of the synapse when it is in its $i$'th state is given by the corresponding element of the column vector $\w$.
We will frequently refer to this function as the memory curve.
It can be thought of as the fraction of synapses that maintain the ideal synaptic strength at time $t$, dictated by the stored memory at time $t=0$.

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode description:
%
\begin{equation}\label{eq:eigendecomp}
  \frg = \sum_a -q_a \uv^a \vv^a,
  \quad
  \vv^a \uv^b = \delta_{ab},
  \quad
  \frg \uv^a = -q_a \uv^a,
  \quad
  \vv^a \frg = -q_a \vv^a.
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
  &\quad \text{where}&\;&
  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
  & \text{and}&&
  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
We can then ask the question: what are the constraints on these quantities, namely initial SNR's $\CI_a$ and time constants $\tau_a$ for the modes,  implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our goal, achieved in \autoref{sec:env}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the entire memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
As a corollary of these results, we also derive an upper-bound on the memory lifetime, or the time at which the SNR drops to a given detection threshold.

\subsection{Initial SNR}\label{sec:initial}

Now we will discuss the SNR at $t=0$:
%
\begin{equation}\label{eq:init}
  \snr(0) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \w.
\end{equation}
%
We will find an upper bound on this quantity for \emph{all} possible models and also find the model that saturates this bound.
As maximizing initial SNR does not place any value on later parts of the memory curve, this will produce models with fast decay that may not be good for memory storage.
Nevertheless, the upper bound will be useful in \autoref{sec:env}.

A useful quantity is the equilibrium probability flux between two disjoint sets of states, $\CA$ and $\CB$:
%
\begin{equation}\label{eq:flux}
  \F_{\CA\CB} = \sum_{i\in\CA} \sum_{j\in\CB} r \eq_i \frg_{ij}.
\end{equation}
%
The initial SNR is closely related to the flux between the states with $\w_i=-1$ and those with $\w_j=+1$:
%
\begin{equation}\label{eq:initflux}
  \snr(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}\,.
\end{equation}
%
This inequality becomes an equality if potentiation never decreases the synaptic weight and depression never increases it, which should be a property of any sensible model.

@@

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
Whilst this model has high initial SNR, it also has very fast decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it easily.
This is not a good design for a synapse, but the resulting bound on the initial SNR will prove useful in \autoref{sec:env}.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\hp\aligntop{\includegraphics[width=0.15\linewidth]{max_init_pot-generated.pdf}}\label{fig:max_init_pot}\hp
  \item\aligntop{\includegraphics[width=0.15\linewidth]{max_init_dep-generated.pdf}}\label{fig:max_init_dep}\hp
  \item\hp\aligntop{\includegraphics[width=0.05\linewidth]{binary_det-generated.pdf}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  }\label{fig:max_init}
\end{figure}

\subsection{Area}\label{sec:area}

Now consider the area under the memory curve:
%
\begin{equation}\label{eq:area}
  A = \int_0^\infty\!\!\dr t \, \snr(t).
\end{equation}
%
We will find an upper bound on this quantity as well as the model that saturates it.
The area is useful as it also provides an upper bound on the lifetime of a memory, $\tau(\epsilon)$, defined as the time at which $\snr(\tau)=\epsilon$ (we often take $\epsilon=1$):
%
\begin{equation}\label{eq:arealife}
  A = \int_0^\infty\!\!\dr t \, \snr(t)
    > \int_0^{\tau(\epsilon)}\!\!\dr t \, \snr(t)
    > \int_0^{\tau(\epsilon)}\!\!\dr t \, \epsilon
    = \epsilon \tau(\epsilon).
\end{equation}
%
However, the area still receives a contribution from the tail of the memory curve, after the SNR has decayed below any useful value.
This means that maximizing SNR will favor models with long tails that will not necessarily be good for storing memories.

One important property of the area is that it has a null ``scaling'' degree of freedom.
If all off-diagonal elements of $\M\potdep$ are multiplied by a constant, with diagonal elements adjusted to keep row-sums fixed, this will have two effects.
First, it will scale up the initial signal creation.
Second, it will scale down time.
Combining these two effects, it will leave the area invariant.

This feature allows us to ignore the lower bound on the diagonal elements, $\M\potdep_{ii}>0$.
If we maximize the area ignoring this constraint, the ``scaling'' mode can be used to satisfy it without changing the area.

We also develop an important organizing principle using the theory of first passage times.
The mean first passage time matrix, $\fptb_{ij}$, is defined as the average time it takes to reach state $j$ for the first time, starting from state $j$.
The diagonal elements are defined to be zero.
Then the quantity
%
\begin{equation}\label{eq:kemenyconst}
  \eta \equiv \sum_j \fptb_{ij}\eq_j
\end{equation}
%
is independent of the starting state $i$.
It is known as Kemeny's constant (see \cite[\S4.4]{kemeny1960finite}).
We can define analogous quantities
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
We can put the states in order of decreasing $\eta^+_i$ or increasing $\eta^-_i$.
Because $\eta^+_i+\eta^-_i=\eta$ is independent of $i$, the two orderings are the same.

With the states in this order, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:max_area}, with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

@@

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
This area is given by
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn\brk{k - \sum_j j\eq_j}$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This result is similar to that found numerically by Barrett and van Rossum \cite{Barrett2008discrete} in a slightly different situation.

The ``sticky'' end states result in very slow decay, but they also make it difficult to encode the memory in the first place.
The memory curve for this model starts off very low, but hovers above zero for a very long time.
This is also not a good design for a synapse, but the resulting bound on the area will also prove useful in \autoref{sec:env}.

>>>>

% -*- TeX -*- -*- US -*-
\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figs/"}}
% >> Only for drafts! <<
%\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%
%additional symbols:
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A memory frontier for complex synapses}
%
% Author List:
%
\author{Subhaneil Lahiri and Surya Ganguli\\
Applied Physics Department, Stanford University, Stanford CA\\
\emaillink{sulahiri@stanford.edu}, \emaillink{sganguli@stanford.edu}
%
}

@@

\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

%Overall, we uncover general design principles governing the functional organization of complex molecular networks, and suggest new experimental observables in synaptic physiology, based on first %passage time theory, that connect molecular complexity to memory.
\end{abstract}

@@

It is widely thought that our very ability to remember the past over long time scales depends crucially on our ability to modify synapses in our brain in an experience dependent manner.
Classical models of synaptic plasticity model synaptic efficacy as an analog scalar value, denoting the size of a post-synaptic potential injected into one neuron from another.
Theoretical work has shown that such models have a reasonable, extensive memory capacity, in which the number of long term associations that can be stored by a neuron is proportional its number of afferent synapses \cite{Hopfield1982model,Amit1985hopfield,Gardner1988perceptron}.
However, recent experimental work has shown that many synapses are more digital than analog; they cannot robustly assume an infinite continuum of analog values, but rather can only take on a finite number of distinguishable strengths, a number than can be as small as two \cite{Bliss1993LTP,Petersen1998allornone,O'Connor2005switch} (though see \cite{Enoki2009graded}).
This one simple modification leads to a catastrophe in memory capacity:  classical models with digital synapses, when operating in a palimpset mode in which the ongoing storage of new memories can overwrite previous memories, have a memory capacity proportional to the logarithm of the number of synapses \cite{amit1992constraints,amit1994learning}.
Intuitively, when synapses are digital, the storage of a new memory can flip a population of synaptic switches, thereby rapidly erasing previous memories stored in the same synaptic population. This result indicates that the dominant theoretical basis for the storage of long term memories in modifiable synaptic switches is flawed.

@@

Theoretical studies of complex, metaplastic synapses, have focused on analyzing the memory performance of a limited number of very specific, molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \autoref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

A simple mean memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ in those synapses that are potentiated and $\eq\M\dep$ in those synapses that are depressed.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{equation}\label{eq:evol}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{equation}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.
Its stationary distribution is $\eq$.

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \e^{rt\frg} \w,
\end{equation}
%
where the weight of the synapse when it is in its $i$'th state is given by the corresponding element of the column vector $\w$.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
  \frg = \sum_a -q_a \uv^a \vv^a,
  \quad
  \vv^a \uv^b = \delta_{ab},
  \quad
  \frg \uv^a = -q_a \uv^a,
  \quad
  \vv^a \frg = -q_a \vv^a.
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
  &\quad \text{where}&\;&
  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
  & \text{and}&&
  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
We can then ask the question: what are the constraints on these quantities, namely initial SNR's $\CI_a$ and time constants $\tau_a$ for the modes,  implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved in \autoref{sec:env}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

\subsection{Bounding initial SNR}\label{sec:initial}

We now give an upper bound on the initial SNR,
%
\begin{equation}\label{eq:init}
  \snr(0) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \w,
\end{equation}
%
over \emph{all} possible models and also find the model that saturates this bound.
A useful quantity is the equilibrium probability flux between two disjoint sets of states, $\CA$ and $\CB$:
%
\begin{equation}\label{eq:flux}
  \F_{\CA\CB} = \sum_{i\in\CA} \sum_{j\in\CB} r \eq_i \frg_{ij}.
\end{equation}
%
The initial SNR is closely related to the flux from the states with $\w_i=-1$ to those with $\w_j=+1$:
%
\begin{equation}\label{eq:initflux}
  \snr(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}\,.
\end{equation}
%
This inequality becomes an equality if potentiation never decreases the synaptic weight and depression never increases it, which should be a property of any sensible model.

@@

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
We note that while this model has high initial SNR, it also has very fast memory decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it rapidly.
This is one example of the tradeoff between optimizing memory at early versus late times.
\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\hp\aligntop{\includegraphics[width=0.15\linewidth]{max_init_pot-generated.pdf}}\label{fig:max_init_pot}\hp
  \item\aligntop{\includegraphics[width=0.15\linewidth]{max_init_dep-generated.pdf}}\label{fig:max_init_dep}\hp
  \item\hp\aligntop{\includegraphics[width=0.05\linewidth]{binary_det-generated.pdf}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  }\label{fig:max_init}
\end{figure}

\subsection{Imposing order on internal states through first passage times}\label{sec:order}

Our goal of understanding the relationship between structure and function in the space of all possible synaptic models is complicated by the fact that this space contains many different possible network topologies, encoded in the nonzero matrix elements of $\M\potdep$.  To systematically analyze this entire space, we develop an important organizing principle using the theory of first passage times in the stochastic process of forgetting, described by $\frg$.
The mean first passage time matrix, $\fptb_{ij}$, is defined as the average time it takes to reach state $j$ for the first time, starting from state $j$.
The diagonal elements are defined to be zero.

A remarkable theorem we will exploit is that the quantity
%
\begin{equation}\label{eq:kemenyconst}
  \eta \equiv \sum_j \fptb_{ij}\eq_j
\end{equation}
%
is independent of the starting state $i$.
It is known as Kemeny's constant (see \cite[\S4.4]{kemeny1960finite}).
Intuitively, \eqref{eq:kemenyconst} states that the average time it takes to reach any state, weighted by its equilibrium probability, is independent of the starting state, implying a hidden constancy inherent in any stochastic process.

In the context of complex synapses, we can define the partial sums
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
Using these definitions, we can then impose an order on the states by arranging them in order of decreasing $\eta^+_i$ or increasing $\eta^-_i$.
Because $\eta^+_i+\eta^-_i=\eta$ is independent of $i$, the two orderings are the same.
In this order, which depends sensitively on the structure of $\M\potdep$, states later (to the right in figures below) can be considered to be more potentiated than states earlier (to the left in figures below), despite the fact that they have the same synaptic efficacy.  
In essence, in this order, a state is considered to be more potentiated if the average time it takes to reach all the strong efficacy states is shorter. 
We will see that synaptic models that optimize various measures of memory have an exceedingly simple structure when, and only when, their states are arranged in this order.

\subsection{Bounding area}\label{sec:area}

Now consider the area under the memory curve:
%
\begin{equation}\label{eq:area}
  A = \int_0^\infty\!\!\dr t \, \snr(t).
\end{equation}
%
We will find an upper bound on this quantity as well as the model that saturates it.

First passage time theory introduced in the \hyperref[sec:order]{previous section} becomes useful because the area has a simple expression in terms of quantities introduced in \eqref{eq:kemenypm}:
%
\begin{equation}\label{eq:areaEta}
\begin{aligned}
  A &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \prn{\M\pot_{ij}-\M\dep_{ij}} \prn{\eta^+_i-\eta^+_j}\\
    &= \sqrt{N}(4f\pot f\dep) \sum_{ij} \eq_i \prn{\M\pot_{ij}-\M\dep_{ij}} \prn{\eta^-_j-\eta^-_i}.
\end{aligned}
\end{equation}
%

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}.
For example, in \autoref{fig:perts}\ref{fig:tri_right}, for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\eta^+_i > \eta^+_j$), we have proven that increasing $\M\pot_{ij}$ and decreasing $\M\dep_{ij}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:max_area}, with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

@@

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$:
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn\brk{k - \sum_j j\eq_j}$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

The ``sticky'' end states result in very slow decay of memory, but they also make it difficult to encode the memory in the first place, since a small fraction of synapses are able to change synaptic efficacy during the storage of a new memory.  
Thus models that optimize area optimize memory at late times, at the expense of early times.

\begin{figure}[tbp]
 \begin{center}
  \includegraphics[width=0.35\linewidth]{multistate_sticky-generated.pdf}
 \end{center}
  \caption{Model that maximizes the area.
  Unlabeled transitions have probability 1.
  Labeled transitions have probability $\varepsilon\to0$.
  }\label{fig:max_area}
\end{figure}
